Contracts for Contracts

### 2023-06-17 paper/contract_solidity.tex
sec 3

We now introduce \lang informally with examples.
\lang is a non-breaking extension that additionally provides means
to specify behavioral contracts on top of the Solidity language.
We often use \emph{specification} for behavioral contracts to avoid terminological ambiguity.

@@

\paragraph{Second-Class Enforcement}
Addresses, as first-class values, can of course escape from a function, e.g.
by returning from the function or storing in external data structures.

Similarly, when a guarded address is used as an argument to another address
call, we lose the control of ...

\paragraph{Second-Class Enforcement}
\note{GW: an easy option is to discard enforcement of escaped addresses at all.
But it seems we can guard escaped addresses within the current contract scope.
We need to think about layered/stacked contracts of addresses (lax/picky).}
Address contracts are \emph{not} higher-order -- if an address takes another
address as argument, then we cannot enforce the contract of the argument
address.
For example, in the following case, \code{b} itself has an ``address contract''
\code{<b-pre> -> <b-post>} but there is no general way to enforce that.
Because we do not have the control of the actual callee function of \code{a},
therefore no way to modify the code of \code{a} to check the precondition
and postcondition when \code{b} is called in \code{a}.

\begin{lstlisting}
  # { a | { b | <b-pre> -> <b-post> } -> { (true, _) | true } }
  function f(address a) {
    address b = ...
    (bool success, bytes memory data) = a.call(b);
  }
\end{lstlisting}

Address contracts are \emph{second-class} -- we can only enforce the
precondition and postcondition of addresses within the current calling
context.  If the address escapes (e.g. by returning, or stored in a global
variable), we cannot enforce the contract anymore.

Why? Because in general we do not have the ability to insert checks around the
computation of that address. We can indeed create a new contract (and a new address) that wraps the old address call with checks, but then calling
the new address will not exhibit the same behavior as calling the old address
for the caller.

For example, the following identity function of addresses has a spec
for the argument address. But the body of function does not invoke
the address, instead, it directly returns \code{a} to the caller.
Once address \code{a} escapes, we cannot enforce the condition anymore.

\begin{lstlisting}
  # { a | { {value, gas}(arg) | <pre-cond> } -> { (res, data) | <post-cond> } }
  function f(address a) returns (address) { return a }
\end{lstlisting}

\paragraph{Notes on Higher-Order Functions}
Functions in Solidity are first-class too, meaning that functions can be used
as arguments for and returned from functions.
However, real closures following lexical scopes do not exist in Solidity
at this moment \footnote{Solidity version 8.20}.
There is no way to write anonymous functions (lambda expressions) or nested,
named functions that capture variables from environments.
Only explicitly defined top-level functions can be used in first-class ways.
This restriction discourages programmers to use higher-order functions due to
its inconvenience and verbosity.
Moreover, it poses challenges to seamlessly implement first-class behavioral
contract and monitor system that guards higher-order functions (e.g. as in
\cite{DBLP:conf/icfp/FindlerF02}) without an expensive whole program
transformation.
Therefore, in this paper we focus on the contract and monitoring of addresses
values and leave monitoring for higher-order functions as future work when
Solidity has a proper support for lambda expressions.

>>>>

We now introduce \lang informally with examples.
\lang is a non-breaking extension that additionally provides means
to specify behavioral contracts on top of the Solidity language.
We often use (executable) \emph{specification} for behavioral contracts to avoid terminological ambiguity.

@@

\paragraph{Second-Class Monitoring}
%\note{GW: an easy option is to discard enforcement of escaped addresses at all.
%But it seems we can guard escaped addresses within the current contract scope.
%We need to think about layered/stacked contracts of addresses (lax/picky).}
Addresses, as first-class values, can of course escape from a function, e.g.
by returning from the function or storing in external data structures.
However, it is not always possible to enforce the specification
of addresses when escaped to third-parties.
For example, consider a function that returns an address equipped with
conditions:
\begin{lstlisting}
EXAMPLE TODO
function f(address a) returns (address) { return a; }
\end{lstlisting}
In this case, if the returned address value crosses the boundary to other
Ethereum programs (by external calls initiated by a third-party) that we do not
have control, we are impotence to monitor if the conditions are violated.

In spite of that, \lang does guarantee that addresses are completely monitored
within the current Ethereum program's calling context, following a stack
discipline.
In other words, for an argument addresses \code{addr} of function \code{f},
\lang is able to detect any violation of \code{addr}'s conditions before
\code{f} returns (even if the call happens in other functions).
We called this mechanism \emph{second-class monitoring}, borrowing Strachey's taxonomy
of first- and second-class objects \cite{DBLP:journals/lisp/Strachey00}.
This is already useful to prevent \note{XXX} \note{fwd ref}, especially when
the address comes from an untrusted party.

Similarly, when a guarded address is used as an argument to another address
call, \note{we lose the control of ...}

\paragraph{Notes on Higher-Order Functions}
Likewise, functions in Solidity can be used as arguments of functions and returned from
functions.
However, they are not really first-class objects, since at this moment
\footnote{Solidity version 8.20} there is no way to write anonymous functions
(lambda expressions) or nested, named functions that capture variables from
environments.
Only explicitly defined top-level functions can be used in ``first-class''
ways.
This restriction prohibits real closure values following lexical scopes,
familiarized by functional programmers.
%This restriction discourages programmers to use higher-order functions due to
%its inconvenience and verbosity.
Moreover, it poses challenges to seamlessly implement first-class behavioral
contract and monitor system that guards higher-order functions (e.g. as in
\cite{DBLP:conf/icfp/FindlerF02}) without an expensive whole program
transformation.
Therefore, in this paper we focus on the contract and monitoring of addresses
values and leave monitoring for higher-order functions as future work when
Solidity has a proper support for lambda expressions.
